\chapter{Methodology}
\label{cha:methodology}
To ensure consistency throughout the description of the methodology, we define the domains and the variables involved in the image-to-image translation and anomaly detection pipeline:
\begin{itemize}
    \item $\mathcal{X}_{sim}$: The domain of \textbf{simulated radargrams}, derived purely from surface topography. A single simulated observation is denoted as $x_{sim} \in \mathcal{X}_{sim}$.
    \item $\mathcal{X}_{real}$: The domain of \textbf{real SHARAD radargrams}, containing surface clutter, thermal noise, and potential subsurface reflectors. A single real observation is denoted as $x_{real} \in \mathcal{X}_{real}$.
    \item $G$: The primary \textbf{Generator} network, parameterized to map the simulated domain to the real domain, $G: \mathcal{X}_{sim} \rightarrow \mathcal{X}_{real}$.
    \item $x_{gen}$: The \textbf{generated background} (or synthetic radargram), produced by the generator from a simulation, such that $x_{gen} = G(x_{sim})$.
    \item $\mathcal{M}$: The resulting \textbf{anomaly map} obtained by comparing the generated baseline $x_{gen}$ with the corresponding real observation $x_{real}$ using a distance metric $\mathcal{D}$.
\end{itemize}

The challenge addressed in this thesis is the discrimination of genuine subsurface geological interfaces from surface clutter and noise in radargrams.
Due to the absence of ground truth labels for subsurface anomalies and the extreme difficulty of obtaining them through in-situ validation, traditional supervised approaches are infeasible.
Instead, we propose an unsupervised anomaly detection framework based on Simulation-to-Real (Sim-to-Real) image-to-image translation.

We train a generative model to translate geometric-optics simulations (containing surface clutter only) into realistic SHARAD radargrams.
This model will systematically fail to reconstruct genuine subsurface reflectors, as these are absent from the simulation domain.
These reconstruction failures manifest as statistically significant residuals in the difference between real observations and model-generated backgrounds, enabling the detection of subsurface anomalies without requiring labeled training data.

This chapter details the complete pipeline implementing this approach.
We first describe the system architecture and data engineering enabling scalable processing of SHARAD datasets (Section~\ref{sec:data_acquisition}).
Sections~\ref{sec:preprocessing_patching}--\ref{sec:modular_implementation} then present data pre-processing, patch extraction, and the design of the generative architectures.
Finally, Section~\ref{sec:anomaly_detection_methodology} describes the residual-based anomaly detection pipeline and the synthetic anomaly injection protocol used for quantitative evaluation.

\input{04_methodology/01_System_Architecture_and_Data_Engineering}
\input{04_methodology/02_The_Dataset_MRO_SHARAD_Radar_Data}
\input{04_methodology/03_Data_Pre-processing_and_Patch_Extraction_Strategy}
\input{04_methodology/04_Resource-Constrained_Data_Loading}
\input{04_methodology/05_Modular_Neural_Network_Implementation}
\input{04_methodology/06_Advanced_Training_Monitoring_and_Visualization}
\input{04_methodology/07_Environment_Isolation_and_Reproducibility}
\input{04_methodology/08_Automated_Hyperparameter_Optimization_Framework}
\input{04_methodology/09_Anomaly_Detection_Methodology}
\input{04_methodology/10_Synthetic_Anomaly_Injection_Strategy}
\input{04_methodology/11_Conclusion}
